Statistical Searches for Microlensing Events in Large, Non-Uniformly Sampled Time-Domain Surveys: A Test Using Palomar Transient Factory Data

### 2013-11-04 text/paper/ms.tex

%\slugcomment{DRAFT \today}
\shorttitle{Microlensing \& PTF}
\shortauthors{Price-Whelan et al.}

\bibliographystyle{apj}

@@

\begin{abstract}
Many current photometric, time-domain surveys are driven by specific goals, such as searches for supernovae or transiting exoplanets, which set the cadence with which individual fields are re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to extremely non-uniform sampling over the survey's nearly $20,000~\mathrm{deg}^2$ footprint. While the median $7.26~\mathrm{deg}^2$ PTF field has been imaged $\sim$40 times in \textit{R}-band, $\sim$$2300~\mathrm{deg}^2$ have been observed more than 100 times. We use the existing PTF data to study the trade-off between searching for microlensing events in a survey whose footprint is much larger than that of typical microlensing searches, but that has far-from-optimal time sampling. To examine the probability that microlensing events can be recovered in these data, we test previous statistics used on uniformly sampled data to identify variables and transients. We find that one such statistic, the von Neumann ratio, also known as the Durbin-Watson statistic, performs best for identifying simulated microlensing events in our data. We develop a selection method using this statistic and apply it to data from all PTF fields with $>$10 $R$-band observations --- a total of $1.1\times10^9$ light curves --- to uncover three candidate sparse-field microlensing events. These candidates cannot be confirmed, as we lack simultaneous, multi-color photometry of the events. However, their number is consistent with predictions for the microlensing event rate in the PTF footprint over the survey's three years of operations, as estimated from near-field microlensing models. This work can help constrain all-sky event rate predictions and tests microlensing signal recovery in large data sets, both of which will be useful to future wide-field, time-domain surveys, such as that planned with the Large Synoptic Survey Telescope. 
\end{abstract}

@@

\section{Introduction}
Over the past 20 years, microlensing, in which gravitational lensing causes a transient increase in the flux from a background point source, has been used to search for dark and compact objects \citep{original_macho, oslowski2008, sartore2010}, to study Galactic structure and kinematics \citep{binney2000}, to determine the shape of stars \citep{rattenbury2005}, and to identify extrasolar planets \citep[][and references therein]{gaudi2011}. Such studies were once limited by the small number of detected events, but thanks to advances in CCD technology and the development of dedicated microlensing surveys, a few thousand events are now observed each year.\footnote{e.g., {\tt http://ogle.astrouw.edu.pl/ogle4/ews/ews.html}} In the coming decade, this number is expected to increase as the next generation of photometric, time-domain surveys comes online, providing an opportunity for the precise study of otherwise hard-to-characterize objects, such as low-mass stars, sub-stellar objects, and isolated neutron stars.

Microlensing surveys typically focus on high-density stellar regions, such as the Galactic Bulge, M31, or the Magellanic Clouds \citep[e.g.,][]{original_ogle, original_macho, eros_original, crotts1996}, but microlensing events are not limited to dense stellar fields. Indeed, lensing events away from these fields are potentially very interesting. While the microlensing event rate in low-density stellar regions will be smaller, compared to events found by modern microlensing surveys, these are expected to involve closer sources and lenses, and less crowded backgrounds \citep{mesolensing}. Thus, such lenses offer more opportunities to constrain the properties of the events.

@@

Only one event outside of a dense stellar field has been recorded to date: the Tago event \citep{fukui2007, gaudi2008}, a serendipitous detection of the microlensing of a bright source. \cite{gaudi2008} set probabilistic limits on the mass, distance, proper motion, and magnitude of this lens using existing observations near the event's location, and further observations may turn these limits into precise measurements. Such an analysis is usually not possible for events in high-density fields where the lenses are, on average, much farther away and blending makes photometry --- and, therefore, mass measurements --- challenging. By contrast, high-Galactic-latitude events probe lenses with distances $\lesssim$1~kpc; the Einstein radii, therefore, tend to be larger ($\sim$milliarcseconds) and may cause detectable astrometric signatures \citep[][]{han2008, mesolensing, mesolensing2}. A measurement of the astrometric shift due to microlensing will break the distance-mass degeneracy in the lensing parameters and enable a direct measurement of the lens mass. Finding even a handful of nearby lenses that can be studied in the same kind of detail as the Tago event would, therefore, be very valuable.

@@

\section{PTF: Survey Design and Consequences for Microlensing Searches}\label{sec:ptf} 
PTF data are collected using the former Canada-France-Hawaii Telescope 12K$\times$8K mosaic camera, which has 11 working chips, $0.92\times10^8$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}, mounted on the 48-nch Oschin Schmidt Telescope (the P48) at Palomar Observatory, CA. Under median seeing conditions (2.2$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey \citep[SDSS;][]{york00} $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in each 60~s exposure \citep{nick2009,rau2009,nick2010}.

@@

\section{Searching for events in the full PTF dataset}\label{sec:search}
To identify candidate microlensing events we: %in the $1.1\times10^9$ light curves with $>$10 $R$-band observations, we:
%We apply the 1\% FPR cut to the 2097 PTF fields ($\sim$15,224 deg$^2$) with $>$10 $R$-band observations using the variability index $\eta$. Our procedure is as follows:
\begin{enumerate}
	\item select the light curves from one of the 2097 fields with $>$10 $R$-band observations;
	\item identify those light curves with more than 50\% good data points and $>$10 good observations (see Section~\ref{sec:detection_eff}) and run Monte Carlo simulations to compute the 1\% FPR limiting values for $\eta$, $\Delta \chi^2$, and $J$ for that field;
	\item compute $\eta$ for all the light curves in the field. While $\sim$10\% of the light curves pass the quality cut described above, only $\sim$0.1-0.5\% of these survive the FPR cut determined from the Monte Carlo simulations;
	\item for fields with SDSS coverage, we use Data Release 9 photometry \citep{dr9paper} to remove sources typed as Galaxy or QSO (using \texttt{objc\_type}\footnote{{\tt http://www.sdss3.org/dr9/algorithms/classify.php}}) and flag candidate quasars \citep[using the cuts described in][]{richards02}; this eliminates $\sim$90\% of the remaining light curves;
	\item fit a microlensing event model and subtract the model, then recompute $\eta$ for all the selected light curves;
	\item reject the light curves for which the new value of $\eta$ still passes the cut\footnote{These are probably periodic variables.} --- this step eliminates $\sim$90-95\% of the remaining light curves;
	\item retain any surviving light curves for further inspection. Typically, there is $\lapprox$1 such light curve per field.
\end{enumerate}
% Turn this into a table, with number of objects left over after each cut?

@@

\acknowledgments
We thank E.~Bowsher, S.~Douglas, and S.~Tang for their assistance in obtaining spectra for our candidates, and M.~Modjaz for help in interpreting several of these spectra. We are grateful to the anonymous referee and to our editor, Eric Feigelson, for comments that improved the paper.

@@

Funding for SDSS-III has been provided by the Alfred P. Sloan Foundation, the Participating Institutions, the National Science Foundation, and the U.S. Department of Energy Office of Science. The SDSS-III web site is http://www.sdss3.org/.

@@

%\bigskip
%\setlength{\baselineskip}{0.6\baselineskip}
\bibliography{references_3}
%\setlength{\baselineskip}{1.667\baselineskip}

>>>>

%\slugcomment{DRAFT \today}
\shorttitle{Microlensing \& PTF}
\shortauthors{Price-Whelan et al.}
\bibliographystyle{apj}

@@

\begin{abstract}
Many current photometric, time-domain surveys are driven by specific goals, such as searches for supernovae or transiting exoplanets, which set the cadence with which individual fields are re-imaged. In the case of the Palomar Transient Factory (PTF), several such sub-surveys are being conducted in parallel, leading to extremely non-uniform sampling over the survey's nearly $20,000~\mathrm{deg}^2$ footprint. While the median $7.26~\mathrm{deg}^2$ PTF field has been imaged $\sim$40 times in \textit{R}-band, $\sim$$2300~\mathrm{deg}^2$ have been observed more than 100 times. We use the existing PTF data to study the trade-off between searching for microlensing events in a survey whose footprint is much larger than that of typical microlensing searches, but that has far-from-optimal time sampling. To examine the probability that microlensing events can be recovered in these data, we test previous statistics used on uniformly sampled data to identify variables and transients. We find that one such statistic, the von Neumann ratio (also known as the Durbin-Watson statistic), performs best for identifying simulated microlensing events in our data. We develop a selection method using this statistic and apply it to data from all PTF fields with $>$10 $R$-band observations --- a total of $1.1\times10^9$ light curves --- to uncover three candidate sparse-field microlensing events. These candidates cannot be confirmed, as we lack simultaneous, multi-color photometry of the events. However, their number is consistent with predictions for the microlensing event rate in the PTF footprint over the survey's three years of operations, as estimated from near-field microlensing models. This work can help constrain all-sky event rate predictions and tests microlensing signal recovery in large data sets, both of which will be useful to future wide-field, time-domain surveys, such as that planned with the Large Synoptic Survey Telescope. 
\end{abstract}

@@

\section{Introduction}
Over the past 20 years, microlensing, in which gravitational lensing causes a transient increase in the flux from a background point source, has been used to search for dark and compact objects \citep{original_macho, oslowski2008, sartore2010}, to study Galactic structure and kinematics \citep{binney2000}, to determine the shape of stars \citep{rattenbury2005}, and to identify extrasolar planets \citep[][and references therein]{gaudi2011}. Such studies were once limited by the small number of detected events, but thanks to advances in CCD technology and the development of dedicated microlensing surveys, a few thousand events are now observed each year.\footnote{e.g., \url{http://ogle.astrouw.edu.pl/ogle4/ews/ews.html}.} In the coming decade, this number is expected to increase as the next generation of photometric, time-domain surveys comes online, providing an opportunity for the precise study of otherwise hard-to-characterize objects, such as low-mass stars, sub-stellar objects, and isolated neutron stars.

Microlensing surveys typically focus on high-density stellar regions, such as the Galactic Bulge, M31, or the Magellanic Clouds \citep[e.g.,][]{original_ogle, original_macho, eros_original, crotts1996}, but microlensing events are not limited to dense stellar fields. Indeed, lensing events away from these fields are potentially very interesting. While the microlensing event rate in low-density stellar regions will be smaller, compared to events found by modern microlensing surveys, these are expected to involve closer sources and lenses, and less crowded backgrounds \citep{mesolensing1}. Thus, such lenses offer more opportunities to constrain the properties of the events.

@@

Only one event outside of a dense stellar field has been recorded to date: the Tago event \citep{fukui2007, gaudi2008}, a serendipitous detection of the microlensing of a bright source. \cite{gaudi2008} set probabilistic limits on the mass, distance, proper motion, and magnitude of this lens using existing observations near the event's location, and further observations may turn these limits into precise measurements. Such an analysis is usually not possible for events in high-density fields where the lenses are, on average, much farther away and blending makes photometry --- and, therefore, mass measurements --- challenging. By contrast, high-Galactic-latitude events probe lenses with distances $\lesssim$1~kpc; the Einstein radii, therefore, tend to be larger ($\sim$milliarcseconds) and may cause detectable astrometric signatures \citep[][]{han2008, mesolensing1, mesolensing2}. A measurement of the astrometric shift due to microlensing will break the distance-mass degeneracy in the lensing parameters and enable a direct measurement of the lens mass. Finding even a handful of nearby lenses that can be studied in the same kind of detail as the Tago event would, therefore, be very valuable.

@@

\section{PTF: Survey Design and Consequences for Microlensing Searches}\label{sec:ptf} 
PTF data are collected using the former Canada-France-Hawaii Telescope 12K$\times$8K mosaic camera, which has 11 working chips, $0.92\times10^8$ pixels, and a 7.26 deg$^2$ field-of-view \citep{rahmer2008}, mounted on the 48-inch Oschin Schmidt Telescope (the P48) at Palomar Observatory, CA. Under median seeing conditions (2.2$\arcsec$), observations in Mould $R$ or Sloan Digital Sky Survey \citep[SDSS;][]{york00} $g$ achieve 2.0$\arcsec$ full-width half-maximum images and reach 5$\sigma$ magnitudes of $R \approx 21.0$ and $g \approx 21.3$ mag in each 60~s exposure \citep{nick2009,rau2009,nick2010}.

@@

\section{Searching for events in the full PTF dataset}\label{sec:search}
To identify candidate microlensing events we: %in the $1.1\times10^9$ light curves with $>$10 $R$-band observations, we:
%We apply the 1\% FPR cut to the 2097 PTF fields ($\sim$15,224 deg$^2$) with $>$10 $R$-band observations using the variability index $\eta$. Our procedure is as follows:
\begin{enumerate}
	\item select the light curves from one of the 2097 fields with $>$10 $R$-band observations;
	\item identify those light curves with more than 50\% good data points and $>$10 good observations (see Section~\ref{sec:detection_eff}) and run Monte Carlo simulations to compute the 1\% FPR limiting values for $\eta$, $\Delta \chi^2$, and $J$ for that field;
	\item compute $\eta$ for all the light curves in the field. While $\sim$10\% of the light curves pass the quality cut described above, only $\sim$0.1-0.5\% of these survive the FPR cut determined from the Monte Carlo simulations;
	\item for fields with SDSS coverage, we use Data Release 9 photometry \citep{dr9paper} to remove sources typed as Galaxy or QSO (using \texttt{objc\_type}\footnote{\url{http://www.sdss3.org/dr9/algorithms/classify.php}}) and flag candidate quasars \citep[using the cuts described in][]{richards02}; this eliminates $\sim$90\% of the remaining light curves;
	\item fit a microlensing event model and subtract the model, then recompute $\eta$ for all the selected light curves;
	\item reject the light curves for which the new value of $\eta$ still passes the cut\footnote{These are probably periodic variables.} --- this step eliminates $\sim$90-95\% of the remaining light curves;
	\item retain any surviving light curves for further inspection. Typically, there is $\lapprox$1 such light curve per field.
\end{enumerate}
% Turn this into a table, with number of objects left over after each cut?

@@

\acknowledgments
We thank E.~Bowsher, S.~Douglas, and S.~Tang for their assistance in obtaining spectra for our candidates, and M.~Modjaz for help in interpreting several of these spectra. We are grateful to the anonymous referee and to our editor, E.~Feigelson, for comments that improved the paper.

@@

Funding for SDSS-III has been provided by the Alfred P. Sloan Foundation, the Participating Institutions, the National Science Foundation, and the U.S. Department of Energy Office of Science. The SDSS-III web site is \url{http://www.sdss3.org/}.

@@

%\bigskip
\setlength{\baselineskip}{0.6\baselineskip}
\bibliography{references_3}
\setlength{\baselineskip}{1.667\baselineskip}
